---
title: "Week 3 Assignment"
author: "Juan Osorio"
date: "24/6/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Overview

This presentation contains a plot of the 500 most populous cities in the world according to the data from [Simple Maps](https://simplemaps.com/data/world-cities). The population is rescaled to millions and it is colored accordingly. The plot is made with the plotly package.

The whole code (including the markdown file) is available on [my github](https://github.com/juanferov/ds_coursera/tree/master/9_ddp/w3).

## Data Processing

Code used for processing the data.

``` {r libraries, results='hide', cache=TRUE}
library(data.table)
library(plotly)
```

```{r processing-code, echo=TRUE, results='hide', cache=TRUE}
cols.keep <- c("city_ascii", "lat", "lng", "country", "population")
dat <- fread("worldcities.csv", select=cols.keep)

dat$population <- dat$population/1e6

dat <- dat[order(-population)]

dat <- dat[1:500,]

dat$popup <- paste(dat$city_ascii, "<br> Country:", dat$country,
                   "<br> Population:", as.character(round(dat$population, 3)),
                   "Millions", sep=" ")
```

## 500 Most Populous Cities

```{r map-plot, fig.align='center', fig.width=8, fig.height=6, cache=TRUE, echo=FALSE, cache=TRUE}
# specify map projection/options
g <- list(
    scope = "world",
    showland = TRUE,
    landcolor = toRGB("grey83"),
    subunitcolor = toRGB("white"),
    countrycolor = toRGB("white"),
    showlakes = TRUE,
    lakecolor = toRGB("white"),
    showsubunits = TRUE,
    showcountries = TRUE
)

# Creating the map
my_map <- plot_geo(dat, lat=~lat, lon=~lng) %>%
    add_markers(text=dat$popup, color=dat$population) %>%
    colorbar(title = "Population in Millions") %>%
    layout(title="Interactive Map", geo=g)

my_map
```

## Thanks


